\documentclass[12pt]{article}

% House style preamble
\input{.house-style/preamble.tex}

\hypersetup{
  pdftitle={Expert Grammaticality Judges as Evaluators, Not Participants},
  pdfkeywords={grammaticality judgments, acceptability judgments, research ethics, human-subjects review, rater reliability}
}

\title{Expert Grammaticality Judges as Evaluators, Not Participants\\
{\large Grammaticality Judgments, Rater Roles, and Research Ethics Review}}
% Anonymized review copy: compile with  xelatex "\def\blind{}\input{main.tex}"
\ifdefined\blind
  \hypersetup{pdfauthor={}}
  \author{\textit{Author name and affiliation withheld for anonymous review}}
  \date{}
\else
  \author{Brett Reynolds \orcidlink{0000-0003-0073-7195}%
  \thanks{Contact: \href{mailto:brett.reynolds@humber.ca}{brett.reynolds@humber.ca}}\\
  Humber Polytechnic \& University of Toronto}
  \date{\today}
\fi

\begin{document}
\maketitle

\begin{abstract}
A grammaticality judgment is always a human act, but it isn't always participant data. Whether the supplier is a participant or an evaluator should turn on what the judgment is evidence of: a person (reactions, processing, or dialect) or an object (an item's status under a stated standard). The role isn't only an ethical classification; it fixes which generalization the judgment licenses, over speakers or items. For the object, the judge resembles an essay scorer or prescriptive annotator more than a survey respondent, and the work's systematic, reliability-checked character marks it as measurement, not a study of the judges. Canada's Tri-Council Policy Statement (TCPS 2) reaches the line by focus of inquiry; the United States Common Rule states it in its \enquote{about whom} definition of a human subject. The proposal routes ethics review rather than loosening it, and neither licenses private intuition nor erases the cases where judges are participants.
\end{abstract}

\noindent\textbf{Keywords:} grammaticality judgments; acceptability judgments; research ethics; human-subjects review; rater reliability

\input{sections/01-problem}
\input{sections/02-role-distinction}
\input{sections/03-essay-evaluator-analogy}
\input{sections/04-linguistic-judgment-work}
\input{sections/05-boundary-cases}
\input{sections/06-regulatory-asymmetry}
\input{sections/07-reporting-and-ethics}
\input{sections/08-conclusion}

\ifdefined\blind\else
\section*{Use of AI tools}

The large language models Claude (Anthropic; Opus 4.8) and GPT-5.5 (OpenAI), used as released and accessed through their providers' web and programming interfaces in June 2026, served as drafting and editing aids. I used them to draft and revise prose, to develop and stress-test the argument and its counterarguments, and to propose candidate sources. Every source, quotation, and citation was verified against the primary document before inclusion; no bibliographic content was taken from model output unverified. I conceived the thesis, performed all verification, and am responsible for all claims, arguments, errors, and interpretive choices. I declare no competing interest arising from the tools' use.
\fi

\clearpage
\printbibliography

\end{document}
