\section{The problem}
\label{sec:problem}

Marking a sentence acceptable is one act with more than one evidential life: a response sampled from a population, a datum about processing or dialect, a consultant's contribution of community knowledge, a colleague's check on a constructed contrast, or an expert's assessment of an item under a stated standard. Ask whether the grammaticality judgments in a linguistics paper needed consent and ethics review, and the answer turns on which life is in play: on whether the study is about that person, not just whether a person produced the judgment.

In Canadian institutions the question is sharp from the start. Canada's Tri-Council Policy Statement (TCPS 2) defines participants through the data and responses relevant to a research question, so on a literal reading a requested grammaticality judgment looks captured before its evidential role has even been described. The argument below is that the role, not the bare request, should settle the classification.

The pull toward the first description has a respectable source. A strand of experimental syntax has argued that informal intuitions, often the author's own, are a weak evidential base, and that acceptability should be measured on samples of naive speakers under controlled conditions \citep{gibsonfedorenko2010,gibsonfedorenko2013,gibson2011mturk}. On that view the design samples the judges, takes their responses as data, and counts as participant research. Informal intuitions can mislead, and the corrective has improved the field.

The corrective answers a measurement question, not a role question. That informal intuitions can be unreliable is a reason to calibrate, sample, or formalize the measurement, and the reliability of informal judgments has itself been tested, with results more favourable than the sharpest critiques predicted \citep{sprouse2013,schutze2016}. It isn't a reason to treat everyone who supplies a grammaticality judgment as the object of study.

The answer isn't that expert judgment escapes scrutiny, but that the scrutiny should follow the role. A judge is a participant when the question concerns that person's reaction, processing, background, identity, or distribution of responses. A judge is an evaluator when the question concerns the linguistic status of an item and expertise is used to assess it against disciplinary standards. Misclassifying costs both ways: treating evaluation as sampling overstates what the judgment shows, and treating sampling as evaluation strips persons of protections they're owed.

The stakes aren't only ethical. The role also fixes which generalization a judgment licenses, over a kind of item or over a population of speakers, so misreading it corrupts the inference, not just the oversight.

Expert judges of grammaticality shouldn't be treated as participants merely because their judgments are human acts. The first question is what role the judgment plays in the evidential design.
