\section{Participants, informants, and evaluators}
\label{sec:role-distinction}

The central distinction is between producing data about oneself and applying expertise to a target object. A participant contributes evidence about a population of people, a psychological process, a distribution of acceptability, a demographic contrast, or an intervention effect. An evaluator contributes an assessment of an object: an essay, a translation, a diagnosis, a code, a performance, or a linguistic example.

Between the two sits informant, less a third role than a recurring hard case. A fieldwork consultant who supplies primary data about their own language is, by the test in the previous paragraph, contributing evidence about a speaker and a community, which places them on the participant or community-contributor side of the contrast, not the essay-scorer side. The term turns slippery because the same person may later be asked to apply a stated criterion to a specified item, which is evaluation. Informant marks where the two roles meet, not a stable category between them.

The category needs special care in documentary, Indigenous, and community-based work. Where a judgment draws its authority from community membership, lived practice, or collective knowledge rather than from a task-specific criterion applied to a defined object, the safer classification is participant or community contributor, not evaluator. TCPS 2 gives this its own chapter, on research involving the First Nations, Inuit, and Métis peoples of Canada, with requirements of engagement and community guidance that the evaluator framing must not bypass \citep[Chapter~9]{tcps2_2022}. The evaluator category shouldn't become a way to relabel community knowledge as expert service.

Calling someone an evaluator is a functional move, not an honorific one. It doesn't make the judgment infallible, unregulated, or methodologically private; it changes the evidential question. The relevant issues become qualification, calibration, independence, adjudication, documentation, and conflict of interest, not sampling, participant burden, demographic representativeness, or the protection of private personal data.

What counts as \term{expertise} here isn't a professional title alone but a role in a design: a judge is an evaluator when selected to apply a stated standard to a linguistic object, so the judgment's evidential value turns on qualification, independence, and task specification, not credentials. The cases form a spectrum, from authorial self-checking and informal consultation to independent expert evaluation and the systematic collection of judgments from a sample. A linguist reporting their own reaction as one of a population isn't an evaluator in this sense; one applying a disciplinary criterion to a specified item may be.

Research ethics didn't invent the contrast. Data annotation in language technology draws the same line, between descriptive labelling, which sets out to capture and model the annotator's own beliefs, and prescriptive labelling, which asks the annotator to apply a fixed scheme to an object \citep{rottger2022annotation}. The descriptive annotator sits nearer the participant pole; the prescriptive annotator, nearer the evaluator pole.

For grammaticality work, the distinction matters because many expert judgments don't estimate speaker-preference distributions. The task is to decide whether a proposed sentence, construction, contrast, or diagnostic behaves as the theory implies. That use of expertise is evaluative.
