\section{The essay-scorer analogy}
\label{sec:essay-evaluator}

Educational and psychometric measurement draws a line that grammaticality research often blurs: the scorers aren't the scored. Essay scorers, oral-proficiency raters, and constructed-response markers can introduce error, bias, drift, and disagreement, and assessment researchers take that seriously. None of it converts the scorers into participants alongside the students. The same holds inside the research process: a colleague or reviewer who judges whether an author's example is acceptable, or whether a claimed contrast holds, is assessing the materials, not thereby enrolling as a subject.

What makes the analogy bite is that this scoring isn't casual. Constructed-response programs train raters against rubrics, double-score responses, monitor severity and drift, adjudicate disagreements, and model rater effects statistically \citep{mccaffrey2022scoring,aera2014standards}. The scoring is about as systematic, multi-rater, and quantitatively scrutinized as human judgment gets. The raters are still part of the measurement apparatus, not its object.

That ordering matters, because the usual worry about grammaticality judgments runs the other way. The intuition is that once judgment collection becomes systematic, with many judges, fixed materials, and reliability statistics, it must be participant research. The essay-scorer case shows the inference backwards. Systematicity, multiple raters, and reliability analysis are signatures of a measurement procedure, not marks of a study about the judges.

The sharpest form of the worry presses on reliability itself. If a study reports how far several judges agree, isn't it then reporting facts about those people? It's reporting a property of the measurement, not a finding about the judges. Inter-rater agreement is what licenses the step from the judges' verdicts to the status of the object, much as a rater facet functions in psychometric models \citep{aera2014standards}. The statistic characterizes the instrument; the object stays what the study is about.

The experimental-syntax worry of \autoref{sec:problem} has a stronger form: role and measurement can't be separated, because an expert judgment is an instrument only as a calibrated sample of experts, and to calibrate it you have to study them. But calibrating an instrument validates a procedure; it doesn't enrol the instrument as a subject. A thermometer checked against a reference doesn't become the object of the reading, and a scoring panel trained and modelled for severity isn't the object of the study when its agreement underwrites a verdict about the item. Experimental syntax already works this way: questionable judgments, \textcite{phillips2010} notes, are caught by colleagues, audiences, and reviewers before they enter the literature, and that scrutiny is adjudication, an evaluator-side control, not a sample drawn to learn about the judges. \textcite{riemer2009} draws a related evidence/prediction line.

The line can move, though, on a principle rather than a stipulation. An aggregate agreement coefficient is used instrumentally, as a premise that licenses the step from the raters' verdicts to the item's status. A model of rater severity by training, dialect, gender, or theoretical orientation makes that variation the conclusion, so the raters become what the study is about. The same numbers can serve either role; what differs is the direction of inference. The clearest case is a design that contrasts expert with naive agreement: the divergence between the two groups is itself the finding \citep{dabrowska2010}, which puts the judges on the participant side (\autoref{sec:boundary-cases}).

The analogy buys neither infallibility nor exemption from scrutiny. It relocates the scrutiny. If raters can be unreliable, the response is to train, blind, rotate, calibrate, adjudicate, report: the standards that govern scoring, not those that govern participant sampling \citep{aera2014standards}. An expert grammaticality judge deciding whether an item behaves as a theory implies is doing the rater's job, and should be held to the rater's standards.

The analogy isn't exact, and the gap cuts in a useful direction. Essay scorers normally judge performances produced independently of them, whereas grammaticality evaluators often judge examples constructed for the argument they support. That raises the need for disclosure, independence, and, where feasible, blinding to the prediction. It changes the controls the role requires, not the role itself.

The essay analogy has a limit worth pressing: a sentence's grammaticality isn't fixed independently of the judge the way an essay's merits are, but constituted by the community's competence and read through the evaluator's own. Expert diamond grading is the closer model. A grader assigns a stone's colour by comparison with master stones under fixed lighting, so colour here isn't response-independent either, keyed to a community-fixed standard and read through trained perception. Yet the grade is about the stone, inter-grader agreement characterizes the procedure rather than the grader, and no one treats the grade as a datum about the grader. A grammaticality verdict has the same shape: a constituted object, a statable criterion, a competent judge as instrument.
