\section{What expert grammaticality judgments do}
\label{sec:linguistic-judgment-work}

Expert grammaticality judgments can serve several roles. They can screen examples, test diagnostic contrasts, adjudicate corpus candidates, check paraphrases, evaluate minimal pairs, or assess whether a constructed item successfully isolates the intended variable. These aren't the same role as estimating how a population would rate the sentence on a Likert scale.

The methodological burden is correspondingly different. An expert-judgment design should say what qualifies the judge, what the judge was asked to evaluate, what information was available, whether judgments were independent, how disagreements were handled, and whether the judgment bears on form, meaning, register, dialect, processing difficulty, or theoretical classification.

This frame carries a precondition: a standard the evaluator can state and be checked against. Where the criterion is articulable, the judge applies it to the item and the verdict is open to challenge. Where the supposed standard is only the theorist's own intuition in other dress, the evaluator role collapses into the private authority this paper declines to defend, since expert and naive judgments can diverge \citep{dabrowska2010}. The label is earned by a statable criterion, not assumed.

A study in this journal shows the precondition at work. Studying resumptive pronouns in English relative clauses, \textcite{lossWicklund2020} begin from a stated generalization in the descriptive literature: a restrictive relative clause requires a gap, not a resumptive pronoun \citep[][1091]{huddleston2002}. Against that criterion, the resumptive \mention{She got a couch at Sears that it was on sale} is marked unacceptable beside the gapped \mention{She got a couch at Sears that was on sale}, even though the resumptive form is attested in speech. That mark evaluates the sentence under a statable rule; it isn't a datum about whoever produced it.

Where the rule runs out, the role changes. For appositive relatives the literature offered only an informal intuition that resumption is better tolerated, with no settled criterion. So \textcite{lossWicklund2020} did what the participant role calls for: they collected acceptability ratings from a sample of speakers and compared the distributions across clause types. The same phenomenon supports an evaluator's verdict where a criterion can be stated and a participant study where the open question is how speakers respond.

The expertise must also match the object. Competence in one variety, register, or theoretical diagnostic doesn't license evaluation across all of them; a syntactician at home with a standard-variety contrast may be a poor judge of a youth register, a minoritized variety, or second-language classroom usage. Where the judgment concerns a socially located pattern, the relevant competence is competence for that pattern, not disciplinary standing in general.

Level discipline matters here. A grammaticality claim, an acceptability-distribution claim, a processing claim, and a social-distribution claim are different claims. Treating every expert judgment as participant data blurs those levels instead of protecting them.

The role also fixes which projectible predicate a judgment confirms, and over what reference class \citep{Goodman1955}. Read as evaluation, marking a sentence confirms a predicate over linguistic items, that structures of this type are grammatical; read as participation, it confirms a predicate over speakers, that people of this variety accept it. A single asterisk feeds a different induction depending on the role. Treating the expert assessment as a sampled response invites bad inference, such as a confidence interval drawn around one expert as though it sampled a population. The role isn't only an ethical classification; it fixes which generalizations are licit.

One might object that grammaticality, grounded in speakers' competence, makes any judgment evidence about persons after all. A sentence's grammaticality does supervene on the community's competence, so the fact is grounded in speakers. But grounding fixes a truthmaker, not a topic. The objection slides twice: from what makes the property obtain, a population idealized, to what bears it, the item; and from the community whose practice constitutes the language to the individual whose verdict reports its status. The evaluator's competence is the route to the item's status, not what the verdict is about.
