\section{What the regulations actually say}
\label{sec:regulatory-asymmetry}

Of the two regulatory frameworks at issue, the Canadian one poses the harder case. TCPS 2 defines participants as individuals whose data or \enquote{responses to interventions, stimuli or questions by the researcher, are relevant to answering the research question(s)} \citep{tcps2_2022}. A rater's judgment answers a question and is plainly relevant to the research question, so on its face the definition takes in the evaluator. The distinction can't rest on the bare fact that a judgment was requested.

It needn't. TCPS 2 allows \enquote{interaction with individuals who are not themselves the focus of the research, in order to obtain information} \citep{tcps2_2022}. The test it sets is the focus of inquiry: whether the person is the object of study, not whether a response was produced.

Its own example is institutional: \enquote{authorized personnel} releasing role-based information about an organization's policies, procedures, or statistics \enquote{are not considered participants} \citep{tcps2_2022}. The Panel sharpens the test, exempting information that staff \enquote{normally provide as part of their work duties} because \enquote{the information is the focus of the research, not the views of the staff member} \citep{pre_scope}.

A grammaticality judgment isn't institutional fact of that kind; it's a cognitive assessment, not a registrar's enrolment figures. But the test doesn't turn on the kind of information, only on the focus of inquiry. An expert asked whether an item meets a stated criterion is interacted with so that the item, not the expert, can be studied: the focus is the item, and the assessor isn't the object.

The same interpretation marks the boundary the argument has to respect. If researchers also ask those people \enquote{to provide personal opinions outside the scope of their job roles}, review is required \citep{pre_scope}. The evaluator role holds only while the judgment is read as a within-competence assessment of the object. Once a judgment is taken as the person's own reaction, idiolect, or attitude, the focus has moved to the person, and the judge becomes a participant (\autoref{sec:boundary-cases}).

This locates the point about method correctly. That \enquote{choice of methodology and/or intent or ability to publish findings} don't determine whether an activity is research \citep{tcps2_2022} blocks the inference from systematicity to participant status, but it doesn't by itself separate participant from evaluator. The focus-of-inquiry test does that work.

United States policy states the same line more explicitly. Under the Common Rule, a human subject is a living individual \enquote{about whom} an investigator obtains data through intervention or interaction, or whose identifiable private information is used \citep{commonrule2018}. The \mention{from}/\mention{about} contrast is built into the definition: a rater's score can be obtained \emph{from} a person without being \emph{about} that person. Where the American text names the distinction, the Canadian framework reaches it through focus of inquiry. The judge evaluates the item; the item doesn't evaluate the judge.
