\section{Reporting and ethics}
\label{sec:reporting-ethics}

The practical proposal is a reporting distinction that follows the role. Participant studies should report recruitment, eligibility, consent, relevant demographics, compensation, exclusion, and the statistical treatment of responses. Evaluation studies should report a different set \citep{aera2014standards}:

\begin{itemize}
  \item the evaluator's qualifications for the standard applied;
  \item the evaluation task and the criterion or diagnostic used;
  \item the materials and context available to the evaluator;
  \item whether the evaluator was blind to the hypothesis;
  \item independence from the author and the project;
  \item how disagreements were adjudicated;
  \item any conflict of interest or relationship to the work;
  \item whether the judgment concerns form, meaning, register, dialect, processing difficulty, or theoretical classification.
\end{itemize}

This routing doesn't weaken ethics review; it aims it. A study that targets persons belongs in participant review with the usual protections; one that uses expert competence to assess linguistic materials belongs under the standards that govern measurement \citep{aera2014standards}.

One protection can't fall through the gap. Classifying a judge as an evaluator settles their status as a source of human-subjects data; it doesn't settle their standing as a contributor of labour. Even where a graduate student pressed to supply judgments is no one's research subject, they're still owed fair acknowledgement, a reasonable burden, and protection against coercion. Those attach to the person as a worker, a matter of labour ethics rather than human-subjects review, a line sharpened in the parallel debate over whether machine-learning crowdworkers are subjects or workers \citep{kaushik2023crowdworkers}. The evaluator framing relocates one set of obligations and leaves the other untouched.

Several questions travel together but shouldn't answer one another: whether the judgment is evidence about the judge (participant status), whether it's reliable (measurement), whether the person was fairly treated, credited, and free from coercion (labour), and whether their identity or judgment will be disclosed (confidentiality and attribution). A negative answer to the first doesn't settle the rest.

The two roles put different pressures on identity, but naming isn't the diagnostic. Participants are normally anonymized because their identities are data-bearing and disclosure can harm them. Evaluators may need attribution, internal traceability, or conflict-of-interest disclosure, because their qualifications and relationships bear on measurement quality, but they too may be anonymized, for independence, blinding, or protection from retaliation. What a study owes is an account of how it handles attribution, anonymity, and conflict of interest, not a rule that reads the role off whether the person is named.

Misclassifying an evaluator as a participant can become a small instance of what \textcite{haggerty2004} calls \term{ethics creep}: the outward expansion of research-ethics review over activities it was never built to cover. A role-test checks the creep well beyond grammaticality, in any field where qualified assessment of materials is read as a sample of the assessors. It checks the opposite error too, refusing the relabeling that would dodge review where persons are the object.

The institutional claim stays narrow. Local boards decide the scope of their own review, and no definition settles policy. The conceptual point, argued in \autoref{sec:regulatory-asymmetry}, is only that the word \enquote{judgment} shouldn't trigger participant review on its own. What should trigger review is whether the design studies persons as persons. The bridge is normative: evidential role is what participant status should track, not something that by itself settles how a board must classify a study.
